---
title: "Model Predicting Exercise Manner"
author: "Anusha Bhat"
date: "Saturday, November 21, 2015"
output: html_document
---

This project uses data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants that performed barbell lifts correctly and incorrectly in 5 different ways. The goal of the project is to predict the manner in which they did the exercise. 

"classe" variable indicating one of 5 class of activities is used as output variable to be predicted. Variables to predict with are selected from the training set to build the model. 

```{r libreq, echo=FALSE, results="hide", message=FALSE, warning=FALSE}

library(knitr)

library(downloader)

library(caret)

library(tools)

library(R.utils)

library(RCurl)

```

## Analysis

### Model Selection

The training data provided is loaded. Since the raw training set consisted for numerous NA values, the dataset is cleaned for columns with sums NA. Then, factor variables are excluded from the dataset, including variables that did not add value, for example, user_name. The remaining 48 predictors are considered for the model.

```{r loadcleantrain}

fileUrl <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", ssl.verifypeer=0L, followlocation=1L)

writeLines(fileUrl,'training.csv')

traindf <- read.csv("training.csv")

traindf <- traindf[ , apply(traindf, 2, function(x) {!sum(is.na(x))})]

traindf <- traindf[ , -grep("kurtosis|skewness|max|min|amplitude", names(traindf), ignore.case = TRUE)]

traindf <- traindf[, - c(traindf$user_name, traindf$cvtd_timestamp, traindf$new_window)]

```

The training set is split into a 70% by 30% train and test data. 

```{r splitds}

inTrain <- createDataPartition(traindf$classe, p = .7, list = FALSE)

trainingSet <- traindf[inTrain,] 

testingSet <- traindf[-inTrain,]

```

The training data is fit with Random Forest to predict the output class. Random Forest is chosen after fitting the training dataset other models (i.e. gbm and decision trees) based on accuracy of model to predict class in testing dataset. Train Control is used in the model to improve computational performance by resampling the training dataset 4 times for cross validation. The accuracy of the model used is quite high shown in the results.

```{r bestModel, cache=TRUE}

trainCntrl <- trainControl(method="cv", number = 4, allowParallel=TRUE)

modFit <- train(classe~., method="rf", data=trainingSet, trControl=trainCntrl)

modFit

```

### Testing Predictions  

The testing dataset created from the raw training data is used to test the rf model. The accuracy of predictions is 99% and the model is significant. The plot shows red points for rows with incorrect predictions when total arm acceleration and total dumbbell acceleration are plotted against each other.

```{r predMod}

pred <- predict(modFit, testingSet)
conf <- confusionMatrix(pred, testingSet$classe)

conf

testingSet$predRight <- pred == testingSet$classe
qplot(total_accel_arm, total_accel_dumbbell, color=predRight, data=testingSet, main="New Predictions")

```

### Out of Sample Error

The out-of-sample error is calculated on the testing dataset extracted from raw training data. The the OOS error is 0.0099

```{r oosErr}

OOSAccuracy <- sum(pred == testingSet$classe)/length(pred)

OOSErr <- 1 - OOSAccuracy

OOSErr

```

## Results

The model is tested on the raw testing data provided and all 20 cases are predicted correctly for class. 

```{r testML, results='hide', message=FALSE, warning=FALSE}

fileUrl2 <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", ssl.verifypeer=0L, followlocation=1L)

writeLines(fileUrl2,'testing.csv')

testdf <- read.csv("testing.csv")

answers <- predict(modFit, testdf)

pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

pml_write_files(answers)

```

